A Simplified Spectral Deferred Correction Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2020-10-17 simplified_sdc_intro/paper.tex

We solve the Euler equations for compressible, reacting flow.  We'll
focus on one-dimensional flow for this description, and the system
appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity, $u$,
specific total energy, $E$, and nuclear
species mass fractions, $X_k$.  The specific total energy relates to
the specific internal energy as $E = e + u^2/2$.  The
corresponding fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\ \rho X_k u \\ \rho u u + p \\  \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state,
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc)
\end{equation}
 The reactive
sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\ \rho \omegadot_k \\ 0 \\ \rho \dot{S}
  \end{array} \right )
\end{equation}
where $\omegadot_k$ is the creation rate for species $k$ and $\dot{S}$
is the energy generation rate per unit mass.

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sq
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{ccccc}
    u & 0 & \rho & 0 & 0 \\
    0 & u & 0    & 0 & 0 \\
    0 & 0 & u    & 1/\rho & 0 \\
    0 & 0 & \Gamma_1 p & u & 0 \\
    0 & 0 & \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $\Gamma_1$ is an adiabatic index, $\Gamma_1 = d\log p/d\log\rho
|_s$.  Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sq$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sq = \Sqhydro + \Rb(q)
\end{equation}
with
\begin{equation}
\Rb(q) = \left ( \begin{array}{c}
     0 \\
     \omegadot_k \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.

>>>>

We solve the Euler equations for compressible, reacting flow.  We'll
focus on one-dimensional flow for this description, and the system
appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity,
$u$, specific total energy, $E$, and nuclear species mass fractions,
$X_k$, and auxiliary composition variables, $a_l$.  The mass fractions
are constrained to sum to 1, $\sum_k X_k = 1$, but no such constraint
exists on the auxiliary variables.  The specific total energy relates
to the specific internal energy as $E = e + u^2/2$.  The corresponding
fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\
   \rho X_k u \\
   \rho \alpha_l u \\
   \rho u u + p \\
   \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state.  For a system
where the composition is completely specified by the mass fractions, $X_k$, the equation
of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}
However, as we'll see shortly, when we use the NSE table, we are using
the results of a much larger network then can be represented by the
mass fractions, and in this case, we rely on the auxiliary composition
variables, $\alpha_l$, to describe the state of the composition, and
our EOS has the form:
\begin{equation}
p = p(\rho, \alpha_l, e)
\end{equation}

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc)
\end{equation}
 The reactive
sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S}
  \end{array} \right )
\end{equation}
where $\omegadot_k$ is the creation rate for species $k$ and $\dot{S}$
is the energy generation rate per unit mass.

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  \alpha_l \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sq
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{cccccc}
    u & 0 & 0 & \rho & 0 & 0 \\
    0 & u & 0 & 0    & 0 & 0 \\
    0 & 0 & u & 0    & 0 & 0 \\
    0 & 0 & 0 & u    & 1/\rho & 0 \\
    0 & 0 & 0 & \Gamma_1 p & u & 0 \\
    0 & 0 & 0 & \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $\Gamma_1$ is an adiabatic index, $\Gamma_1 = d\log p/d\log\rho
|_s$.  Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sq$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sq = \Sqhydro + \Rb(q)
\end{equation}
with
\begin{equation}
\Rb(q) = \left ( \begin{array}{c}
     0 \\
     \omegadot_k \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.
